Exploring the ages of rotating stars using galactic dynamics: a novel approach to calibrating gyrochronology

### 2020-03-11 paper/conclusion.tex
rocio;s comments
hyphen to en-dash

In this paper, we demonstrated that the dispersion of velocities in the
direction of galactic latitude, \vb, can be used as an age proxy, by showing
that there is no strong evidence for mass-dependent heating in low-mass
\kepler\ dwarfs: the velocity dispersions of K and M dwarfs, whose
main-sequence lifetimes are longer than around 11 Gyrs, do not appear to
increase with decreasing mass.
Although {\it vertical} velocity, \vz, is a quantity that has been
demonstrated to trace time-dependent orbital heating in the disc of the
Galaxy, most stars with measured rotation periods do not yet have radial
velocities, so we used velocity in the direction of Galactic latitude, \vb\,
as a proxy for \vz.
Using stars in the GUMS simulation, we showed that using \vb\ as a proxy for
\vz\ introduces an additional velocity dispersion, which increases with
increasing Galactic latitude.
For this reason we did not attempt to convert \vb\ dispersions into ages using
an age-velocity dispersion relation.
However, after removing high-latitude ($b>15^\circ$) stars from the sample, we
confirmed that using \vb\ instead of \vz\ does not introduce any
mass-dependent velocity dispersion bias into the sample.
We therefore assumed that \vb\ velocity dispersion can be used to accurately
rank stars by age, \ie\ a group of stars with a large velocity dispersion is,
on average, older than a group of stars with a small velocity dispersion.

% We found that old groups of cool dwarfs, selected to be coeval using the
% \citet{angus2019} gyrochronology relation, do {\it not} have the same velocity
% dispersion across all temperatures.
We used the \vb\ velocity dispersions of stars in the \mct\ catalog to explore
the evolution of stellar rotation period as a function of effective
temperature and age.
We found that the \citet{angus2019} relation, which is based on the
period-color relation of the 650 Myr Praesepe cluster, does not correctly
describe the period-age-\teff\ relation for old stars.
Instead we found that, at young ages, rotation period is anti-correlated with
\teff: cooler stars spin more slowly than hotter stars of the same age.
However, at intermediate ages the relation flattens out and K dwarfs rotate at
the same rate, regardless of mass.
At old ages, it seems that cooler K dwarfs spin more rapidly than hotter K
dwarfs of the same age.
We showed that the period-\teff\ relations change shape over time in a way
that qualitatively agrees with theoretical models which include a
mass-dependent core-envelope angular momentum transport \citep{spada2019}.

@@

% We outlined a number of scenarios which could provide alternative
% explanations for these observations, including incorrect dust corrections for
% the lowest-mass stars and an excess of companions increasing the velocity
% dispersion for these stars.

>>>>

In this paper, we demonstrated that the dispersion of velocities in the
direction of galactic latitude, \vb, can be used as an age proxy, by showing
that there is no strong evidence for mass-dependent heating in low-mass
\kepler\ dwarfs: the velocity dispersions of K and M dwarfs, whose
main-sequence lifetimes are longer than around 11 Gyrs, do not appear to
increase with decreasing mass.
Although {\it vertical} velocity, \vz, is a quantity that has been
demonstrated to trace time-dependent orbital heating in the disc of the
Galaxy, most stars with measured rotation periods do not yet have radial
velocities, so we used velocity in the direction of Galactic latitude, \vb\,
as a proxy for \vz.
Using stars in the GUMS simulation, we showed that using \vb\ as a proxy for
\vz\ introduces an additional pseudo-velocity dispersion, which increases with
increasing Galactic latitude.
For this reason we did not attempt to convert \vb\ dispersions into ages using
an age-velocity dispersion relation.
However, after removing high-latitude ($b>15^\circ$) stars from the sample, we
confirmed that using \vb\ instead of \vz\ does not introduce any
mass-dependent velocity dispersion bias into the sample.
We therefore assumed that \vb\ velocity dispersion can be used to accurately
rank stars by age, \ie\ a group of stars with a large velocity dispersion is,
on average, older than a group of stars with a small velocity dispersion.

% We found that old groups of cool dwarfs, selected to be coeval using the
% \citet{angus2019} gyrochronology relation, do {\it not} have the same velocity
% dispersion across all temperatures.
We used the \vb\ velocity dispersions of stars in the \mct\ catalog to explore
the evolution of stellar rotation period as a function of effective
temperature and age.
We found that the \citet{angus2019} relation, which is based on the
period--color relation of the 650 Myr Praesepe cluster, does not correctly
describe the period--age--\teff\ relation for old stars.
Instead we found that, at young ages, rotation period is anti-correlated with
\teff: cooler stars spin more slowly than hotter stars of the same age.
However, at intermediate ages the relation flattens out and K dwarfs rotate at
the same rate, regardless of mass.
At old ages, it seems that cooler K dwarfs spin more rapidly than hotter K
dwarfs of the same age.
We showed that the period--\teff\ relations change shape over time in a way
that qualitatively agrees with theoretical models which include a
mass-dependent core-envelope angular momentum transport \citep{spada2019}.

@@

% explanations for these observations, including incorrect dust corrections for
% the lowest-mass stars and an excess of companions increasing the velocity
% dispersion for these stars.
